---
title: "Intro to ggplot2"
output:
  pdf_document:
    toc: yes
    toc_depth: 3
  html_document:
    highlight: pygments
    theme: cerulean
    toc: yes
    toc_depth: 3
    toc_float: yes
fontsize: 12 pt
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(message = FALSE)
knitr::opts_chunk$set(warning = FALSE) #to supress warning messages in all output
```

***

```{r}
acitelli <- read.csv("acitelli.csv")
```

There are quite a few ways to make figures in R, we'll the popular package `ggplot2`. You can find a cheat sheet for `ggplot2` [here](http://github.com/rstudio/cheatsheets/raw/master/data-visualization-2.1.pdf). Be sure to install it first if you never have, or if you need to update it. 

```{r}
#install.packages("ggplot2")
library(ggplot2)
```

# Visualizing Data with `qplot()`

## Quick Histogram

First, let's make a histogram for `satisfaction`. The easiest way to make a figure with `ggplot2` is with the `qplot()` function. This stands for *quick plot*.  Notice in the code below that we did not specify anything about a histogram. `qplot()` guesses which type of plot we want based on the variable's type (i.e., integer, number, double, factor, character).  

```{r}
qplot(satisfaction, data = acitelli)
```

There are too many bins (it defaults to 30 bins), we can ask for a specific number by adding the `bins = ` argument. Try playing around the bin number below to find the optimal plot. I put `10` in there as a placeholder.

```{r}
#play around with different numbers of bins
qplot(satisfaction, data = acitelli, bins = 10)
```

Remember that you can run the `?` code to see the help file for any function. 

```{r}
?qplot
```

We might want to see if the distributions are different for men and women. We can do this by mapping `Gender` to the fill aesthetic. Note that we could used `color =` if we want a hollow histogram. Try it out!

```{r}
qplot(x = satisfaction, fill = as.factor(gender), data = acitelli, bins = 10)
```

Why did we have to change the variable type of gender? What happens if you don't? 

## Quick Density Plot

An alternative to the histogram is the density plot. It displays a smoothed distribution and the area under the curve always sums to 1, thus, it's good for comparing two groups with different n's. 

```{r}
qplot(x = satisfaction, fill = as.factor(gender), data = acitelli, alpha = .05, geom = "density")
```

What would you say the `alpha = ` argument is doing? What happens if you take it away? What if you change the value?

```{r}
#play with alpha here
```

## Quick Scatterplot

We might also want a scatter plot. Again,`qplot()` guesses what we want, but it's a good idea to specify which variable goes on the x-axis and which goes on the y-axis. 

```{r}
qplot(x = tension, y = satisfaction, data = acitelli)
```

We can even add a third variable, mapping it to color. 

```{r}
qplot(x = tension, y = satisfaction, color = gender, data = acitelli)
```

What's going on with the color scale for gender? Can you explain why this happend? What happens if you change the variable type of gender?

```{r}

```

## Quick Boxplots

We can ask for side-by-side boxplots when our x variable is categorical. In this case `qplot()` does **NOT** know what to do, so we tell it we want boxplots by adding the argument `geom = "boxplot"`. 

```{r}
qplot(y = satisfaction, x = as.factor(gender), data = acitelli, geom = "boxplot")
```

What would happen if you added `fill = as.factor(gender)` as an argument. Try it by adding a new chunk below. 

**ADD A NEW CHUNK HERE**

# Full Blown ggplots with `ggplot()`

## Histogram

For more complex figures we will need to move away from using the `qplot()` function in favor of the heavy duty `ggplot()` function. To get a sense of how `ggplot()` builds plots, first we will just run the empty function.

```{r}
ggplot()
```

Next, we can add the data and start mapping variables to aesthetics. 

```{r}
ggplot(acitelli, aes(x = satisfaction))
```

After we have specified aesthetic mappings, we can then add geoms. Notice that we make use of the `+` symbol with ggplots. The `+` needs to be on the right of each piece of the plot. We add a histogram with the `geom_histogram()` function. 

```{r}
ggplot(acitelli, aes(x = satisfaction)) +
  geom_histogram()
```

Note in the histogram above the y-axis are the counts of observations in each bin. There were some calculations involved in getting these counts. Counts are the default *statistic* when you ask for a histogram. We can change the number of bins by adding `bins =` inside of the `geom_histogram()` function. 

```{r}
ggplot(acitelli, aes(x = satisfaction)) +
  geom_histogram(bins = 12)
```

Just as before, if we want overlayed histograms, we can map gender to the fill aesthetic. 

```{r}
ggplot(acitelli, aes(x = satisfaction)) +
  geom_histogram(aes(fill = as.factor(gender)), bins = 12)
```

Alternatively, we can ask for separate facets for each level of the Gender variable with the `fact_wrap()` function. Notice that there is a `~` before Gender inside of this function. 

```{r}
ggplot(acitelli, aes(x = satisfaction)) +
  geom_histogram(bins = 12) +
  facet_wrap(~as.factor(gender))
```

## Scatter Plot

Next we'll make that scatter plot again. We'll map `tension` to the x-axis and `satisfaction` to the y-axis. Then we'll add `geom_point()`.

```{r}
ggplot(acitelli, aes(x = tension, y = satisfaction)) +
  geom_point()
```

Why does it appear as though there is far less data than there really is? Check out the plot when we use `geom_jitter()`. What do you think `geom_jitter()` does?

```{r}
ggplot(acitelli, aes(x = tension, y = satisfaction)) +
  geom_jitter(height = 0, width = .05, alpha = 0.5)
```

We can add more than one geom. To the jittered scatter plot we can add a least squares regression line with `geom_smooth()`. Inside of geom smooth we need to specific `method = "lm"`, the lm stand for *linear model*. We can also turn off the standard errors with se = 0.  

```{r}
ggplot(acitelli, aes(x = tension, y = satisfaction)) +
  geom_jitter(height = 0, width = .05, alpha = 0.5) +
  geom_smooth(method = "lm", se = 0)
```

Again, we can map gender to the color aesthetic. Note: Here the mapping of gender to color is done in the `geom_smooth()` function---but what would happen if you moved it to the `aes()` function in the `ggplot()` function?

```{r}
ggplot(acitelli, aes(x = tension, y = satisfaction)) +
  geom_jitter() +
  geom_smooth(aes(color = as.factor(gender)), method = "lm", se = 0)
```

Or use `facet_wrap()`. In the code below, what does `color = "salmon"` do? Could you change the regression lines to salmon?

```{r}
ggplot(acitelli, aes(x = tension, y = satisfaction)) +
  geom_jitter(color = "salmon") +
  geom_smooth(method = "lm", se = 0) +
  facet_wrap(~as.factor(gender))
```

## Labels and Colors

We can create a plot object with the `<-` symbol. Then to print the plot we'd need to run a line with the name of our plot. 

```{r}
myplot <- ggplot(acitelli, aes(x = tension, 
                               y = satisfaction, 
                               color = as.factor(gender))) +
  geom_jitter(height = 0, width = .05) +
  geom_smooth(method = "lm", se = 0)

myplot
```

Then, we can add to that plot object. We can add x labels, x labels, change the colors, and the theme. There is much more that you can do with `ggplot2`!

```{r}
myplot + 
  xlab("Tension") +
  ylab("Satisfaction") +
  scale_color_manual(values = c("gold", "dodgerblue"), 
                     name = "Gender") +
  ylim(1,4) +
  xlim(1,4) +
  theme_minimal() 
```

